---
title: "aryw_new"
author: "Kushagra Gupta"
date: "6/6/2021"
output: html_document
---

```{r}
set.seed(10)
library(Rcpp)
library(rbenchmark)
```
#### Replicating eureka.f with RCpp
The RCpp translation of eureka.f is in eureka_check.cpp

```{r}
sourceCpp('eureka_check.cpp')
sourceCpp('eureka.cpp')
source('hard_task.R') # to make a Markov chain
```
ar.yw.new function which is the same as the ar.yw call with the fortran call to eureka.f replaced by a call to the RCpp function eureka().

```{r}
ar.yw.new <-
  function (x, aic = TRUE, order.max = NULL, na.action = na.fail,
            demean = TRUE, series = NULL, ...)
  {
    if(is.null(series)) series <- deparse1(substitute(x))
    ists <- is.ts(x)
    x <- na.action(as.ts(x))
    if(ists) xtsp <- tsp(x)
    xfreq <- frequency(x)
    x <- as.matrix(x)
    if(!is.numeric(x))
      stop("'x' must be numeric")
    if(any(is.na(x) != is.na(x[,1]))) stop("NAs in 'x' must be the same row-wise")
    nser <- ncol(x)
    if (demean) {
      xm <- colMeans(x, na.rm=TRUE)
      x <- sweep(x, 2L, xm, check.margin=FALSE)
    } else xm <- rep.int(0, nser)
    n.used <- nrow(x)
    n.obs <- sum(!is.na(x[,1])) # number of non-missing rows
    order.max <- if (is.null(order.max))
      min(n.obs - 1L, floor(10 * log10(n.obs))) else floor(order.max)
    if (order.max < 1L) stop("'order.max' must be >= 1")
    else if (order.max >= n.obs) stop("'order.max' must be < 'n.obs'")
    xacf <- acf(x, type = "covariance", lag.max = order.max, plot = FALSE,
                demean=demean, na.action = na.pass)$acf
    if(nser > 1L) {
      ## multivariate case
      snames <- colnames(x)
      A <- B <- array(0, dim = c(order.max + 1L, nser, nser))
      A[1L, , ] <- B[1L, , ] <- diag(nser)
      EA <- EB <- xacf[1L, , , drop = TRUE]
      partialacf <- array(dim = c(order.max, nser, nser))
      xaic <- numeric(order.max + 1L)
      solve.yw <- function(m) {
        # Solve Yule-Walker equations with Whittle's
        # generalization of the Levinson(-Durbin) algorithm
        betaA <- betaB <- 0
        for (i in 0L:m) {
          betaA <- betaA + A[i + 1L, , ] %*% xacf[m + 2L - i, , ]
          betaB <- betaB + B[i + 1L, , ] %*% t(xacf[m + 2L - i, , ])
        }
        KA <- -t(qr.solve(t(EB), t(betaA)))
        KB <- -t(qr.solve(t(EA), t(betaB)))
        EB <<- (diag(nser) - KB %*% KA) %*% EB
        EA <<- (diag(nser) - KA %*% KB) %*% EA
        Aold <- A
        Bold <- B
        for (i in seq_len(m + 1L)) {
          A[i + 1L, , ] <<- Aold[i + 1L, , ] + KA %*% Bold[m + 2L - i, , ]
          B[i + 1L, , ] <<- Bold[i + 1L, , ] + KB %*% Aold[m + 2L - i, , ]
        }
      }
      cal.aic <- function(m) { # (EA)  omits mean params, that is constant adj
        logdet <- determinant.matrix(EA)$modulus
        # == log(abs(prod(diag(qr(EA)$qr))))
        n.obs * logdet + 2 * m * nser * nser
      }
      cal.resid <- function() {
        resid <- array(0, dim = c(n.used - order, nser))
        for (i in 0L:order)
          resid <- resid +
            tcrossprod(x[(order - i + 1L):(n.used - i), , drop = FALSE],
                       ar[i + 1L, , ])
        rbind(matrix(NA, order, nser), resid)
      }
      order <- 0L
      for (m in 0L:order.max) {
        xaic[m + 1L] <- cal.aic(m) # (EA)
        if (!aic || xaic[m + 1L] == min(xaic[seq_len(m + 1L)])) {
          ar <- A
          order <- m
          var.pred <- EA * n.obs/(n.obs - nser * (m + 1L))
        }
        if (m < order.max) {
          solve.yw(m) #-> update (EA, EB, A, B)
          partialacf[m + 1L, , ] <- -A[m + 2L, , ]
        }
      }
      xaic <- setNames(xaic - min(xaic), 0L:order.max)
      resid <- cal.resid()
      if(order) {
        ar <- -ar[2L:(order + 1L), , , drop = FALSE]
        dimnames(ar) <- list(seq_len(order), snames, snames)
      } else ar <- array(0, dim = c(0L, nser, nser),
                         dimnames = list(NULL, snames, snames))
      dimnames(var.pred) <- list(snames, snames)
      dimnames(partialacf) <- list(seq_len(order.max), snames, snames)
      colnames(resid) <- colnames(x)
    } else { ## univariate case
      if (xacf[1L] == 0) stop("zero-variance series")
      r <- as.double(drop(xacf))
      #print("yoohoo")
      z <- eureka_check(as.integer(order.max), r, r, matrix(double(order.max^2), nrow = order.max), 
                  double(order.max), double(order.max))
      # z <- .Fortran(C_eureka, as.integer(order.max), r, r,
      #               coefs = double(order.max^2),
      #               vars = double(order.max),
      #               double(order.max))
      coefs <- matrix(z$coefs, order.max, order.max)
      partialacf <- array(diag(coefs), dim = c(order.max, 1L, 1L))
      var.pred <- c(r[1L], z$vars)
      xaic <- n.obs * log(var.pred) + 2 * (0L:order.max) + 2 * demean
      maic <- min(aic)
      xaic <- setNames(if(is.finite(maic)) xaic - min(xaic) else
        ifelse(xaic == maic, 0, Inf),
        0L:order.max)
      order <- if (aic) (0L:order.max)[xaic == 0L] else order.max
      ar <- if (order) coefs[order, seq_len(order)] else numeric()
      var.pred <- var.pred[order + 1L]
      ## Splus compatibility fix
      var.pred <- var.pred * n.obs/(n.obs - (order + 1L))
      resid <- if(order) c(rep.int(NA, order), embed(x, order + 1L) %*% c(1, -ar))
      else as.vector(x) # we had as.matrix() above
      if(ists) {
        attr(resid, "tsp") <- xtsp
        attr(resid, "class") <- "ts"
      }
    }
    res <- list(order = order, ar = ar, var.pred = var.pred, x.mean  =  drop(xm),
                aic  =  xaic, n.used = n.used, n.obs = n.obs, order.max = order.max,
                partialacf = partialacf, resid = resid, method = "Yule-Walker",
                series = series, frequency = xfreq, call = match.call())
    if(nser == 1L && order)
      res$asy.var.coef <- var.pred/n.obs *
      solve(toeplitz(drop(xacf)[seq_len(order)]))
    class(res) <- "ar"
    res
  }
```
```{r}
ans = RWMH(1e5, 0, 0.05) 
ans = drop(ans) # 1D Markov chain

foo1 = ar(ans, aic = TRUE, method = "yule-walker")
foo2 = ar.yw.new(ans, aic = TRUE)
print(foo1$ar)
print(foo2$ar)

if(foo1$ar == foo2$ar)
  print("same coefficients")
```
Time comparision

```{r}
for(i in 1:10)  {
  x = rnorm(1e5)
  foo = benchmark(replications = 10, ar(x, aic = TRUE, method = "yule-walker"), ar.yw.new(x, aic = TRUE))
  print(foo)
  ar.fit = ar(x, aic = TRUE, method = "yule-walker")
  print(ar.fit$order)
}

```
```{r}
benchmark(replications = 1e2, ar(ans, aic = TRUE, method = "yule-walker"), ar.yw.new(ans, aic = TRUE))
```

The methods are more or less comparable in terms of speed, with the fortran code maybe edging slightly.

#### Adding a threshold

The Levinson algorithm used for calculating the AR coefficients runs recursively, i.e. it uses coefficients obtained from fitting AR(p-1) to obtain coefficients from fitting AR(p). The current fortran implementation estimates coefficients for fitting AR(p) for p: 1 to order.max irrespective of the value of aic parameter to ar(). We have added a check to the eureka implementation which compares the coefficient to a threshold. If at any time a coefficient falls below the threshold, we don't compute coefficients beyond that order (i.e. we don't always complete the loop from 1 to order.max) and return. Therefore, we have added a threshold check in the Levinson algorithm which can stop the recurssion going from 1 to order.max if at any time any estimated coefficient falls below the threshold.

```{r}
ar.yw.threshold <-
  function (x, aic = TRUE, order.max = NULL, na.action = na.fail,
            demean = TRUE, series = NULL, ...)
  {
    if(is.null(series)) series <- deparse1(substitute(x))
    ists <- is.ts(x)
    x <- na.action(as.ts(x))
    if(ists) xtsp <- tsp(x)
    xfreq <- frequency(x)
    x <- as.matrix(x)
    if(!is.numeric(x))
      stop("'x' must be numeric")
    if(any(is.na(x) != is.na(x[,1]))) stop("NAs in 'x' must be the same row-wise")
    nser <- ncol(x)
    if (demean) {
      xm <- colMeans(x, na.rm=TRUE)
      x <- sweep(x, 2L, xm, check.margin=FALSE)
    } else xm <- rep.int(0, nser)
    n.used <- nrow(x)
    n.obs <- sum(!is.na(x[,1])) # number of non-missing rows
    order.max <- if (is.null(order.max))
      min(n.obs - 1L, floor(10 * log10(n.obs))) else floor(order.max)
    if (order.max < 1L) stop("'order.max' must be >= 1")
    else if (order.max >= n.obs) stop("'order.max' must be < 'n.obs'")
    xacf <- acf(x, type = "covariance", lag.max = order.max, plot = FALSE,
                demean=demean, na.action = na.pass)$acf
    if(nser > 1L) {
      ## multivariate case
      snames <- colnames(x)
      A <- B <- array(0, dim = c(order.max + 1L, nser, nser))
      A[1L, , ] <- B[1L, , ] <- diag(nser)
      EA <- EB <- xacf[1L, , , drop = TRUE]
      partialacf <- array(dim = c(order.max, nser, nser))
      xaic <- numeric(order.max + 1L)
      solve.yw <- function(m) {
        # Solve Yule-Walker equations with Whittle's
        # generalization of the Levinson(-Durbin) algorithm
        betaA <- betaB <- 0
        for (i in 0L:m) {
          betaA <- betaA + A[i + 1L, , ] %*% xacf[m + 2L - i, , ]
          betaB <- betaB + B[i + 1L, , ] %*% t(xacf[m + 2L - i, , ])
        }
        KA <- -t(qr.solve(t(EB), t(betaA)))
        KB <- -t(qr.solve(t(EA), t(betaB)))
        EB <<- (diag(nser) - KB %*% KA) %*% EB
        EA <<- (diag(nser) - KA %*% KB) %*% EA
        Aold <- A
        Bold <- B
        for (i in seq_len(m + 1L)) {
          A[i + 1L, , ] <<- Aold[i + 1L, , ] + KA %*% Bold[m + 2L - i, , ]
          B[i + 1L, , ] <<- Bold[i + 1L, , ] + KB %*% Aold[m + 2L - i, , ]
        }
      }
      cal.aic <- function(m) { # (EA)  omits mean params, that is constant adj
        logdet <- determinant.matrix(EA)$modulus
        # == log(abs(prod(diag(qr(EA)$qr))))
        n.obs * logdet + 2 * m * nser * nser
      }
      cal.resid <- function() {
        resid <- array(0, dim = c(n.used - order, nser))
        for (i in 0L:order)
          resid <- resid +
            tcrossprod(x[(order - i + 1L):(n.used - i), , drop = FALSE],
                       ar[i + 1L, , ])
        rbind(matrix(NA, order, nser), resid)
      }
      order <- 0L
      for (m in 0L:order.max) {
        xaic[m + 1L] <- cal.aic(m) # (EA)
        if (!aic || xaic[m + 1L] == min(xaic[seq_len(m + 1L)])) {
          ar <- A
          order <- m
          var.pred <- EA * n.obs/(n.obs - nser * (m + 1L))
        }
        if (m < order.max) {
          solve.yw(m) #-> update (EA, EB, A, B)
          partialacf[m + 1L, , ] <- -A[m + 2L, , ]
        }
      }
      xaic <- setNames(xaic - min(xaic), 0L:order.max)
      resid <- cal.resid()
      if(order) {
        ar <- -ar[2L:(order + 1L), , , drop = FALSE]
        dimnames(ar) <- list(seq_len(order), snames, snames)
      } else ar <- array(0, dim = c(0L, nser, nser),
                         dimnames = list(NULL, snames, snames))
      dimnames(var.pred) <- list(snames, snames)
      dimnames(partialacf) <- list(seq_len(order.max), snames, snames)
      colnames(resid) <- colnames(x)
    } else { ## univariate case
      if (xacf[1L] == 0) stop("zero-variance series")
      r <- as.double(drop(xacf))
      #print("yoohoo")
      z <- eureka(as.integer(order.max), r, r, matrix(double(order.max^2), nrow = order.max), 
                  double(order.max), double(order.max), 0.01)
      # z <- .Fortran(C_eureka, as.integer(order.max), r, r,
      #               coefs = double(order.max^2),
      #               vars = double(order.max),
      #               double(order.max))
      order.max = z$order
      if(z$order) {
        coefs <- matrix(z$coefs[1:order.max, 1:order.max], nrow = order.max)
        #coefs <- matrix(z$coefs, order.max, order.max)
        partialacf <- array(diag(coefs), dim = c(order.max, 1L, 1L))
        var.pred <- c(r[1L], z$vars[1:order.max])
        #var.pred <- c(r[1L], z$vars)
        xaic <- n.obs * log(var.pred) + 2 * (0L:order.max) + 2 * demean
        maic <- min(aic)
        xaic <- setNames(if(is.finite(maic)) xaic - min(xaic) else
          ifelse(xaic == maic, 0, Inf),
          0L:order.max)
        order <- if (aic) (0L:order.max)[xaic == 0L] else order.max
        ar <- if (order) coefs[order, seq_len(order)] else numeric()
        var.pred <- var.pred[order + 1L]
        ## Splus compatibility fix
        var.pred <- var.pred * n.obs/(n.obs - (order + 1L))
      }
      else  {
        order = 0
        ar = 0
        var.pred = 0
        xaic = 0
        partialacf = 0
      }
      
      resid <- if(order) c(rep.int(NA, order), embed(x, order + 1L) %*% c(1, -ar))
      else as.vector(x) # we had as.matrix() above
      if(ists) {
        attr(resid, "tsp") <- xtsp
        attr(resid, "class") <- "ts"
      }
    }
    res <- list(order = order, ar = ar, var.pred = var.pred, x.mean  =  drop(xm),
                aic  =  xaic, n.used = n.used, n.obs = n.obs, order.max = order.max,
                partialacf = partialacf, resid = resid, method = "Yule-Walker",
                series = series, frequency = xfreq, call = match.call())
    if(nser == 1L && order)
      res$asy.var.coef <- var.pred/n.obs *
      solve(toeplitz(drop(xacf)[seq_len(order)]))
    class(res) <- "ar"
    res
  }
```

Time comparision

```{r}
for(i in 1:10)  {
  x = rnorm(1e5)
  foo = benchmark(replications = 10, ar(x, aic = TRUE, method = "yule-walker"), ar.yw.threshold(x, aic = TRUE))
  print(foo)
  ar.fit = ar(x, aic = TRUE, method = "yule-walker")
  ar.fit.thresh = ar.yw.threshold(x, aic = TRUE)
  print(c(ar.fit$order, ar.fit.thresh$order))
}
```

```{r}
benchmark(replications = 1e2, ar(ans, aic = TRUE, method = "yule-walker"), ar.yw.threshold(ans, aic = TRUE))
ar.fit = ar(ans, aic = TRUE, method = "yule-walker")
ar.fit.thresh = ar.yw.threshold(ans, aic = TRUE)
print(c(ar.fit$order, ar.fit.thresh$order))
```
Our custom implementation slightly edges the existing implementation in terms of time. 



